---
title: "To adjust, or not to adjust: that is the question"
author: "Tom Palmer, `tom.palmer@lancaster.ac.uk`, Department of Mathematics and Statistics, Lancaster University"
#date: "MSc Statistics Open Day, 27^th^ February 2019"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 2
runtime: shiny
---

```{r setup, include=FALSE}
library(DiagrammeR)
```

## Introduction

In this exercise data has been simulated under several different scenarios.

In each scenario there are 3 variables, $X$, $C$, and $Y$, and the relationship between these variables is different.

By observing what happens when we fit unadjusted and adjusted linear models to the data you are asked to think about the relationships between the variables.

## Background

In these scenarios the unadjusted model is a model of the form, for $i = 1,\dots,N$ observations
$$
y_i = \beta_0 + \beta_1x_i + \varepsilon_i,\ \varepsilon_i \sim N(0, \sigma^2).
$$

The adjusted model includes $C$ as an additional additive covariate, i.e.
$$
y_i = \beta_0 + \beta_1x_i + \beta_2c_i + \varepsilon_i,\ \varepsilon_i \sim N(0, \sigma^2).
$$

## What you need to do

- In the tab for each scenario there are two buttons to press to fit the unadjusted and adjusted models.
- Click both buttons and look at the estimates of the coefficients.
- The data has been simulated such that the true value of the estimated coefficient on $X$ should be 1.
- After fitting both models think about the estimates that you have found:
  + In which model have you obtained a coefficient on $X$ close to 1?
  + Write down (or remember) what you think the relationship between the variables might be.
- To reveal the answer click the third button.

## Scenarios{.tabset .tabset-fade .tabset-pills}

### Scenario 1

```{r, include=FALSE}
simindpred <- function(seed = 12345){
  set.seed(seed)
  n <- 1000
  c <- rnorm(n)
  x <- rnorm(n)
  y <- x + c + rnorm(n)
  data.frame(x, c, y)
}
```

```{r, echo=FALSE}
shinyApp(

  ui = fluidPage(
    actionButton(inputId = "click1", label = "Click to run unadjusted model"),
    verbatimTextOutput("fit1"),
    
    actionButton(inputId = "click2", label = "Click to run adjusted model"),
    verbatimTextOutput("fit2"),
    
    actionButton(inputId = "click3", label = "Click to reveal description"),
    grVizOutput("dag1"),
    textOutput("textout1")
    
  ),

  server = function(input, output) {
    dat <- simindpred()
    
    observeEvent(input$click1, {
      output$fit1 = renderPrint({
        summary(lm(y ~ x, data = dat))
      })      
    })
    
    observeEvent(input$click2, {
      output$fit2 = renderPrint({
        summary(lm(y ~ x + c, data = dat))
      })      
    })
    
    observeEvent(input$click3, {
      output$dag1 = renderGrViz({
        DiagrammeR::grViz("
      digraph dag {

      graph [rankdir=TB]

      node [shape=plaintext]
      C
      X
      Y

      { rank = same; X Y }

      C -> Y
      X -> Y [minlen=2]
      }
      ", height = 100)
      })
    })
    
    observeEvent(input$click3, {
      output$textout1 <- renderText({
        "In this example X and C are independent predictors of Y."
        })
    })
    
  },

  options = list(height = 1000)
)
```

### Scenario 2

```{r, include=FALSE}
simconf <- function(seed = 12345){
  set.seed(seed)
  n <- 1000
  c <- rnorm(n)
  x <- c + rnorm(n)
  y <- x + c + rnorm(n)
  data.frame(x, c, y)
}
```

```{r, echo=FALSE}
shinyApp(

  ui = fluidPage(
    actionButton(inputId = "click1", label = "Click to run unadjusted model"),
    verbatimTextOutput("fit1"),
    
    actionButton(inputId = "click2", label = "Click to run adjusted model"),
    verbatimTextOutput("fit2"),
    
    actionButton(inputId = "click3", label = "Click to reveal description"),
    grVizOutput("dag1"),
    textOutput("textout1")
    
  ),

  server = function(input, output) {
    dat <- simconf()
    
    observeEvent(input$click1, {
      output$fit1 = renderPrint({
        summary(lm(y ~ x, data = dat))
      })      
    })
    
    observeEvent(input$click2, {
      output$fit2 = renderPrint({
        summary(lm(y ~ x + c, data = dat))
      })      
    })
    
    observeEvent(input$click3, {
      output$dag1 = renderGrViz({
        DiagrammeR::grViz("
      digraph dag {

      graph [rankdir=TB]

      node [shape=plaintext]
      C
      X
      Y

      { rank = same; X Y }

      C -> X
      C -> Y
      X -> Y [minlen=2]
      }
      ", height = 100)
      })
    })
    
    observeEvent(input$click3, {
      output$textout1 <- renderText({
        "This example demonstrates confounding."
        })
    })
    
  },

  options = list(height = 1000)
)
```

### Scenario 3

```{r, include=FALSE}
simmed <- function(seed = 12345){
  set.seed(seed)
  n <- 1000
  x <- rnorm(n)
  c <- x + rnorm(n)
  y <- x + c + rnorm(n)
  data.frame(x, c, y)
}
```

```{r, echo=FALSE}
shinyApp(

  ui = fluidPage(
    actionButton(inputId = "click1", label = "Click to run unadjusted model"),
    verbatimTextOutput("fit1"),
    
    actionButton(inputId = "click2", label = "Click to run adjusted model"),
    verbatimTextOutput("fit2"),
    
    actionButton(inputId = "click3", label = "Click to reveal description"),
    grVizOutput("dag1"),
    textOutput("textout1")
    
  ),

  server = function(input, output) {
    dat <- simmed()
    
    observeEvent(input$click1, {
      output$fit1 = renderPrint({
        summary(lm(y ~ x, data = dat))
      })      
    })
    
    observeEvent(input$click2, {
      output$fit2 = renderPrint({
        summary(lm(y ~ x + c, data = dat))
      })      
    })
    
    observeEvent(input$click3, {
      output$dag1 = renderGrViz({
        DiagrammeR::grViz("
      digraph dag {

      graph [rankdir=TB]

      node [shape=circle]
      X [label='X']
      C [label='C']
      Y [label='Y']

      { rank = same; X C Y }

      X -> C [minlen=2]
      C -> Y [minlen=2]
      X -> Y [minlen=4]
      }
      ", height = 100)
      })
    })
    
    observeEvent(input$click3, {
      output$textout1 <- renderText({
        "This example demonstrates mediation, with C being the mediator because it lies on the causal pathway between X and Y."
        })
    })
    
  },

  options = list(height = 1000)
)
```

### Scenario 4

```{r, include=FALSE}
simcoll <- function(seed = 12345){
  set.seed(seed)
  n <- 1000
  x <- rnorm(n)
  y <- x + rnorm(n)
  c <- x + y + rnorm(n)
  data.frame(x, c, y)
}
```

```{r, echo=FALSE}
shinyApp(

  ui = fluidPage(
    actionButton(inputId = "click1", label = "Click to run unadjusted model"),
    verbatimTextOutput("fit1"),
    
    actionButton(inputId = "click2", label = "Click to run adjusted model"),
    verbatimTextOutput("fit2"),
    
    actionButton(inputId = "click3", label = "Click to reveal description"),
    grVizOutput("dag1"),
    textOutput("textout1")
    
  ),

  server = function(input, output) {
    dat <- simcoll()
    
    observeEvent(input$click1, {
      output$fit1 = renderPrint({
        summary(lm(y ~ x, data = dat))
      })      
    })
    
    observeEvent(input$click2, {
      output$fit2 = renderPrint({
        summary(lm(y ~ x + c, data = dat))
      })      
    })
    
    observeEvent(input$click3, {
      output$dag1 = renderGrViz({
        DiagrammeR::grViz("
      digraph dag {

      graph [rankdir=TB]

      node [shape=circle]
      X [label='X']
      C [label='C']
      Y [label='Y']

      { rank = same; X Y }

      X -> Y [minlen=3]
      X -> C
      Y -> C [minlen=.5]
      }
      ", height = 100)
      })
    })
    
    observeEvent(input$click3, {
      output$textout1 <- renderText({
        "This example demonstrates a collider, C."
        })
    })
    
  },

  options = list(height = 1000)
)
```

### Scenario 5

```{r, include=FALSE}
siminteraction <- function(seed = 12345){
  set.seed(seed)
  n <- 1000
  c <- rnorm(n)
  x <- rnorm(n)
  y <- x + 5 * c + 10 * x * c + rnorm(n)
  data.frame(x, c, y)
}
```

```{r, echo=FALSE}
shinyApp(

  ui = fluidPage(
    actionButton(inputId = "click1", label = "Click to run unadjusted model"),
    verbatimTextOutput("fit1"),
    
    actionButton(inputId = "click2", label = "Click to run adjusted model"),
    verbatimTextOutput("fit2"),
    
    actionButton(inputId = "click3", label = "Click to reveal description"),
    grVizOutput("dag1"),
    textOutput("textout1"),
    
    actionButton(inputId = "click2a", label = "Click to run third model"),
    verbatimTextOutput("fit2a")
  ),

  server = function(input, output) {
    dat <- siminteraction()
    
    observeEvent(input$click1, {
      output$fit1 = renderPrint({
        summary(lm(y ~ x, data = dat))
      })      
    })
    
    observeEvent(input$click2, {
      output$fit2 = renderPrint({
        summary(lm(y ~ x + c, data = dat))
      })      
    })
    
    observeEvent(input$click3, {
      output$dag1 = renderGrViz({
        DiagrammeR::grViz("
      digraph dag {

      graph [rankdir=TB]

      node [shape=circle]
      C [label='C']
      X [label='X']
      Y [label='Y']

      { rank = same; X Y }

      C -> Y
      X -> Y [minlen=3]
      }
      ", height = 100)
      })
    })
    
    observeEvent(input$click3, {
      output$textout1 <- renderText({
        "In this example the data generating model for Y includes the main effects of X and C and their interaction."
        })
    })
    
    observeEvent(input$click2a, {
      output$fit2a = renderPrint({
        summary(lm(y ~ x * c, data = dat))
      })      
    })

  },

  options = list(height = 1000)
)
```

### Scenario 6

```{r, include=FALSE}
simsquared <- function(seed = 12345){
  set.seed(seed)
  n <- 1000
  c <- rnorm(n)
  x <- rnorm(n)
  y <- x + 5 * x^2 + 5 * c + rnorm(n)
  data.frame(x, c, y)
}
```

```{r, echo=FALSE}
shinyApp(

  ui = fluidPage(
    actionButton(inputId = "click1", label = "Click to run unadjusted model"),
    verbatimTextOutput("fit1"),
    
    actionButton(inputId = "click2", label = "Click to run adjusted model"),
    verbatimTextOutput("fit2"),
    
    actionButton(inputId = "click3", label = "Click to reveal description"),
    grVizOutput("dag1"),
    textOutput("textout1"),
    
    actionButton(inputId = "click2a", label = "Click to run third model"),
    verbatimTextOutput("fit2a")
  ),

  server = function(input, output) {
    dat <- simsquared()
    
    observeEvent(input$click1, {
      output$fit1 = renderPrint({
        summary(lm(y ~ x, data = dat))
      })      
    })
    
    observeEvent(input$click2, {
      output$fit2 = renderPrint({
        summary(lm(y ~ x + c, data = dat))
      })      
    })
    
    observeEvent(input$click3, {
      output$dag1 = renderGrViz({
        DiagrammeR::grViz("
      digraph dag {

      graph [rankdir=TB]

      node [shape=circle]
      C [label='C']
      X [label='X']
      Y [label='Y']

      { rank = same; X Y }

      C -> Y
      X -> Y [minlen=3]
      }
      ", height = 100)
      })
    })
    
    observeEvent(input$click3, {
      output$textout1 <- renderText({
        "In this example the data generating model for Y includes the main effects of X and C and X^2."
        })
    })
    
    observeEvent(input$click2a, {
      output$fit2a = renderPrint({
        summary(lm(y ~ x + I(x^2) + c, data = dat))
      })      
    })

  },

  options = list(height = 1000)
)
```

### Scenario 7

```{r, include=FALSE}
simmbias <- function(seed = 12345){
  set.seed(seed)
  n <- 1000
  a <- rnorm(n)
  b <- rnorm(n)
  c <- a + b + rnorm(n)
  x <- a + rnorm(n)
  y <- x + b + rnorm(n)
  data.frame(x, a, b, c, y)
}
```

```{r, echo=FALSE}
shinyApp(

  ui = fluidPage(
    actionButton(inputId = "click1", label = "Click to run unadjusted model"),
    verbatimTextOutput("fit1"),
    
    actionButton(inputId = "click2", label = "Click to run adjusted model"),
    verbatimTextOutput("fit2"),
    
    actionButton(inputId = "click3", label = "Click to reveal description"),
    grVizOutput("dag1"),
    textOutput("textout1"),
    
    actionButton(inputId = "click2a", label = "Click to run third model"),
    verbatimTextOutput("fit2a")
  ),

  server = function(input, output) {
    dat <- simmbias()
    
    observeEvent(input$click1, {
      output$fit1 = renderPrint({
        summary(lm(y ~ x, data = dat))
      })      
    })
    
    observeEvent(input$click2, {
      output$fit2 = renderPrint({
        summary(lm(y ~ x + c, data = dat))
      })      
    })
    
    observeEvent(input$click3, {
      output$dag1 = renderGrViz({
        DiagrammeR::grViz("
      digraph dag {

      graph [rankdir=TB]

      node [shape=plaintext]
      A [pos='0,2!']
      B [pos='2,2!']
      C [pos='1,1!']
      X [pos='0,0!']
      Y [pos='2,0!']

      { rank = same; A B }
      { rank = same; X Y }

      A -> C 
      B -> C 
      A -> X [minlen=2]
      B -> Y [minlen=2]
      X -> Y [minlen=2]
      }
      ", height = 100)
      })
    })
    
    observeEvent(input$click3, {
      output$textout1 <- renderText({
        "This example uses data generated under the M-bias structure."
        })
    })
    
    observeEvent(input$click2a, {
      output$fit2a = renderPrint({
        summary(lm(y ~ x + a + b + c, data = dat))
      })      
    })

  },

  options = list(height = 1000)
)
```
